--- 
title: "Practical edgeR" 
author: "Malte Thodberg" 
date: "`r Sys.Date()`" 
output: 
  ioslides_presentation: 
     highlight: tango 
     smaller: yes 
footer: Shenzen 2016 
fontsize: 9pt 
vignette: > 
  %\VignetteIndexEntry{Practical PCA} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

## Teachers and slides

- __Malte Thodberg__ 
- PhD Student ~ 1 year 
- Focus: Clinical Bioinformatics 
- Mail: `malte.thodberg@bio.ku.dk` 
- Background: BSc Molecular Biomedicine & MSc Bioinformatics

## Practical Outline

The plan:

- Introduction to `model.matrix` for building model matrices
- Example analysis on a single dataset
- Play around with the remaining datasets.

## Datasets

Example datasets:

- `iris` dataset from `datasets` (default R-package). 
- `oliveoil` dataset from the `pdfCluster` package.

Real datasets:

- `zebrafish` dataset 
- `fission` dataset 
- `pasilla` dataset

Advanced dataset for playing around:

- `tissues` dataset

# Design and model matrices 

## R is a statistical programming language

R has a powerful built-in language for linear models, making it simple to form a model matrix (0-1) representation of a design matrix.

This is done by using the `model.matrix` function with the build in formula language (preceded by `~`).

Let's make a simple dummy design we can explore

Remember, this is how the design matrix for the `zebrafish` dataset looks:

```{r}
dummy <- data.frame(factor1=c("A", "A", "B", "B", "A", "A", "B", "B"),
										 factor2=c("C", "C", "C", "C", "D", "D", "D", "D"))

```

## Basic model

Simple 2-factor design build using `~`:
```{r}
model.matrix(~factor1, data=dummy)
```

## Adding more groups

3-factor design using `+`:
```{r}
model.matrix(~factor1+factor2, data=dummy)
```

## Interaction terms

An an interaction term using `:`:
```{r}
model.matrix(~factor1+factor2+factor1:factor2, data=dummy)
```

## Interaction terms

`*` is shorthand for both base levels and interactions:
```{r}
model.matrix(~factor1*factor2, data=dummy)
```

## Formula cheat sheet.

The R formula interface is almost a programming language in itself!

Due to it's expressiveness, it's used by huge number of packages, making it well worth the effort to learn.

See this cheatsheet for an overview:

https://ww2.coastal.edu/kingw/statistics/R-tutorials/formulae.html

# First steps with edgeR

## Trimming and normalizing the data.

Just as before, we started by loading, trimming and normalizing the data:

```{r}
# Load the data
library(Shenzen2016)
library(ggplot2)
library(edgeR)
data("zebrafish")

# Trim 
EM_zebra <- subset(zebrafish$Expression, rowSums(zebrafish$Expression >= 5) >= 3)

# Calculate normalization factors
dge_zebra <- DGEList(EM_zebra)
dge_zebra <- calcNormFactors(dge_zebra, method="TMM")
```

## Building the model matrix

Before we can estimate dispersions, we must decide on our model matrix. Recall the simpel design of the `zebrafish` dataset:

```{r}
zebrafish$Design
```

We use `model.matrix` to make a simple model matrix:

```{r}
mod <- model.matrix(~gallein, data=zebrafish$Design)
```

## Estimating the dispersion

Now we move on to estimating dispersion or BCV:

```{r}
disp_zebra <- estimateDisp(dge_zebra, design=mod, robust=TRUE)
```

The 3-step estimation is all done by a single function in the newest version of edgeR: `estimateDisp`, which replaces the three previous functions of 

- `estimateCommonDisp`
- `estimateTrendedDisp`
- `estimateTagwiseDisp`

## Plotting the dispersion

We can then plot the estimates:
```{r}
plotBCV(y=disp_zebra)
```

## Fiting gene-wise GLMs and testing a coefficient.

Now we can fit gene-wise GLMs:

```{r}
fit_zebra <- glmFit(disp_zebra, design=mod)
```

Finally, we can test a coefficient in the model for DE:

```{r}
# Using the name of the coefficient
gallein <- glmLRT(fit_zebra, coef="galleintreated")

# Using the index of the coefficient
gallein <- glmLRT(fit_zebra, coef=2)
```

## Inspecting the results

edgeR has several useful functions for inspecting and plotting the results:

Inspecting the top hits with `topTags`:

```{r}
topTags(gallein)
```

## Inspecting the results

Summarize the number of up- and down-regulated genes with `decideTestsDGE`:

```{r}
is_de <- decideTestsDGE(gallein, p.value=0.1)
head(is_de)
summary(is_de)
```

## Inspecting the results

MA-plot, showing relation between mean expression and logFC, :
```{r}
plotSmear(gallein, de.tags=rownames(gallein)[is_de != 0])
```

## Inspecting the results

There is no built-in function for making a volcano plot, but it's easy to do yourself:
```{r}
res <- as.data.frame(topTags(gallein, n=Inf, sort.by="none"))
ggplot(res, aes(x=logFC, y=-log10(PValue), color=FDR < 0.05)) + geom_point()
```

## Quasi-likelihood: Fitting models

Recently, edgeR introduced an alternative DE pipeline using Quasi-likelihood. It should be slightly more conservative than the standard edgeR pipeline. Code-wise, though, they are almost identical:

```{r}
fitQL <- glmQLFit(y=disp_zebra, design=mod, robust=TRUE)
galleinQL <- glmQLFTest(fitQL, coef="galleintreated")
summary(decideTestsDGE(galleinQL, p.value=0.1))
```

## Quasi-likehood: Plotting dispersions

The QL-pipeline estimates slightly different dispersions:

```{r}
plotQLDisp(fitQL)
```

## Final exercises:

Use the above code as a template for conducting your own analysis. For the `yeast` and `pasilla` datasets:

- Load, trim and normalize the data
- Inspect the study design, and decide on a model matrix to use with edgeR
- Use edgeR to estimate dispersions and fit gene-wise GLMs
- Report the number of up- and down-regulated genes for the tests of interests
- Generate smear and volcano plots summarizing the analysis.

If you have time:

- Investigate the effect of different normalization methods.
- Try out the QL-pipeline by replacing `glmFit` with `glmQLFit` and `glmLRT` with `glmQLFTest`
- Refering the slides, see if you can also test a contrast by using a contrast-vector of the form: `c(0,0,-1,1)`

## 
